---
title: "Assignment 2: Coding Basics"
author: "Bryce O'Brien"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on coding basics. 

## Directions
1. Change "Student Name" on line 3 (above) with your name.
2. Work through the steps, **creating code and output** that fulfill each instruction.
3. Be sure to **answer the questions** in this assignment document.
4. When you have completed the assignment, **Knit** the text and code into a single PDF file.
5. After Knitting, submit the completed exercise (PDF file) to the dropbox in Sakai. Add your first and last name into the file name (e.g., "FirstLast_A02_CodingBasics.Rmd") prior to submission.



## Basics Day 1

1. Generate a sequence of numbers from one to 100, increasing by fours. Assign this sequence a name. 

2. Compute the mean and median of this sequence. 

3. Ask R to determine whether the mean is greater than the median. 

4. Insert comments in your code to describe what you are doing.

```{r}
#1. 
lab2_sequence<-seq(1, 100, 4)

#2. 
mean(lab2_sequence)
median(lab2_sequence)


#3. 
recipe_comparison <- function(x) {
  if(mean(x) < median(x)) {
    "Greater Median"
  } 
  else if (mean(x) > median(x)) {
    "Greater Mean"
  }
  else {
    "Mean and Median are equal"
  }
}
view_comparison<-recipe_comparison(lab2_sequence); view_comparison
```

## Basics Day 2

5. Create a series of vectors, each with four components, consisting of (a) names of students, (b) test scores out of a total 100 points, and (c) whether or not they have passed the test (TRUE or FALSE) with a passing grade of 50.

6. Label each vector with a comment on what type of vector it is.

7. Combine each of the vectors into a data frame. Assign the data frame an informative name.

8. Label the columns of your data frame with informative titles.


```{r}
names_of_students<-c("Roxy", "Meg", "Lucy", "Burke")
#names_of_students is a character vector
test_scores<-c(45, 60, 65, 80)
#test_scores is a numerical vector
passed_test<-c(FALSE, TRUE, TRUE, TRUE)
#passed_test is a logical vector

test_info_df<-data.frame("Student.Names"=names_of_students,"Test.Scores"=test_scores,"Passed.Test"=passed_test)

```

9. QUESTION: How is this data frame different from a matrix?

> Answer: 
A data frame can include many different vector types (in this case, our data frame is comprised of numerical, character and logical vectors). A matrix, on the other hand, can only include one vector type (all columns must have the same mode and length).

10. Create a function with an if/else statement. Your function should determine whether a test score is a passing grade of 50 or above (TRUE or FALSE). You will need to choose either the `if` and `else` statements or the `ifelse` statement. Hint: Use `print`, not `return`. The name of your function should be informative.

11. Apply your function to the vector with test scores that you created in number 5. 

```{r}
recipe_passing<- function(x){
  ifelse(test_scores>50, TRUE, FALSE) 
  
}

view_passing<-recipe_passing(test_scores)
view_passing
```

12. QUESTION: Which option of `if` and `else` vs. `ifelse` worked? Why?

> Answer: The function with an 'ifelse' statement worked because the "test_scores" vector has more than one element. Initially, I coded a function with both 'if' and 'else' statements and only the first test score in my column was analyzed (I received a warning message saying: the condition has length > 1 and only the first element will be use).
