---
title: "Assignment 3: Data Exploration"
author: "Bryce O'Brien, Section 2"
output: pdf_document
geometry: margin=2.54cm
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on Data Exploration. 

## Directions
1. Change "Student Name, Section #" on line 3 (above) with your name and section number.
2. Work through the steps, **creating code and output** that fulfill each instruction.
3. Be sure to **answer the questions** in this assignment document.
4. When you have completed the assignment, **Knit** the text and code into a single PDF file.
5. After Knitting, submit the completed exercise (PDF file) to the dropbox in Sakai. Add your last name into the file name (e.g., "FirstLast_A03_DataExploration.Rmd") prior to submission.

The completed exercise is due on <>.


## Set up your R session

1. Check your working directory, load necessary packages (tidyverse), and upload two datasets: the ECOTOX neonicotinoid dataset (ECOTOX_Neonicotinoids_Insects_raw.csv) and the Niwot Ridge NEON dataset for litter and woody debris (NEON_NIWO_Litter_massdata_2018-08_raw.csv). Name these datasets "Neonics" and "Litter", respectively. **Be sure to add the `stringsAsFactors = TRUE` parameter to the function when reading in the CSV files.**
```{r, message = FALSE}
setwd("~/Desktop/Bryce's Things/Environmental_Data_Analytics_2022")
```

```{r,  message = FALSE}
getwd()
library(tidyverse)
Litter<-read.csv("./Data/Raw/NEON_NIWO_Litter_massdata_2018-08_raw.csv", stringsAsFactors = TRUE)
Neonics<-read.csv("./Data/Raw/ECOTOX_Neonicotinoids_Insects_raw.csv", stringsAsFactors = TRUE)
```

## Learn about your system

2. The neonicotinoid dataset was collected from the Environmental Protection Agency's ECOTOX Knowledgebase, a database for ecotoxicology research. Neonicotinoids are a class of insecticides used widely in agriculture. The dataset that has been pulled includes all studies published on insects. Why might we be interested in the ecotoxicologoy of neonicotinoids on insects? Feel free to do a brief internet search if you feel you need more background information. 

> Answer: Many insects provide important ecosystem functions that are necessary for agricultural success (many crops, for example, need to be pollinated). Although Neonicotinoids serve as an effective means of pest control (which is also important for agricultural success), potential negative effects of Neonicotinoids on non-target insects such as pollinators could have cascading ecological and agricicultural implications that should be studied and considered in policy desicions. 

3. The Niwot Ridge litter and woody debris dataset was collected from the National Ecological Observatory Network, which collectively includes 81 aquatic and terrestrial sites across 20 ecoclimatic domains. 32 of these sites sample forest litter and woody debris, and we will focus on the Niwot Ridge long-term ecological research (LTER) station in Colorado. Why might we be interested in studying litter and woody debris that falls to the ground in forests? Feel free to do a brief internet search if you feel you need more background information. 

> Answer: We may be interested in studying litter and woody debris on the forest floor in order to estimate the annual Aboveground Net Primary Productvity (ANPP) and the aboveground biomass at various scales (plot, site, continent). A study of this nature could also provides us with a better unsderstanding of temporal vegetative carbon fluxes. 

4. How is litter and woody debris sampled as part of the NEON network? Read the NEON_Litterfall_UserGuide.pdf document to learn more. List three pieces of salient information about the sampling methods here: 

> Answer: Litter and woody debris sampling is conducted within Terrestrial NEON sites that contain woody vegetation greater than 2m in height. Mass data for each collection event are measured seperately for eight functional groups (leaves, needles, twigs/branches, woody material seeds, flowers, other and mixed). Ground traps are measured once a year and elevated traps are measured on a plot-specific timeline depending on the species of vegetation present. Three pieces of salient information pertainign to sampling methods are as follows:
*Litter is defined as material dropped from the forest canopy that has a butt end diameter of <2cm and a length of <50cm; woody debris is defined as material dropped from the forest canopy that has a butt end diameter of <2cm and a length of >50cm.
*an elevated trap is comprised of 0.5m^2 square holding a mesh basket about 80cm off the forest floor and a ground trap is a 3m x 5m rectangular area. 
*One elevated trap and one ground trap are deployed as a pair for every 400m^2 plot area (reulting in about 1-4 trap pairs per plot).

## Obtain basic summaries of your data (Neonics)

5. What are the dimensions of the dataset?

```{r}
dim(Neonics)
#The Neonics dataset has 4,623 rows and 30 columns
```

6. Using the `summary` function on the "Effect" column, determine the most common effects that are studied. Why might these effects specifically be of interest?
```{r}
summary(Neonics$Effect)
```

> Answer: Population is the most common effect studied in this dataset; population data is particularly informative because it represents broader effects on a species or taxonomic group occupying the same area at a given time.

7. Using the `summary` function, determine the six most commonly studied species in the dataset (common name). What do these species have in common, and why might they be of interest over other insects? Feel free to do a brief internet search for more information if needed.

```{r}
summary(Neonics$Species.Common.Name)
```

> Answer: The six most commonly studied species in this dataset are the Honey Bee, the Parasitic Wasp, the Buff Tailed Bumblebee, the Carniolan Honey Bee, the Bumble Bee and the Italian Honeybee. These species are all pollinators, which means they are paticularly important for agriculture!

8. Concentrations are always a numeric value. What is the class of Conc.1..Author. in the dataset, and why is it not numeric?

```{r}
class(Neonics$Conc.1..Author.)
```

> Answer: Conc.1.Author is a factor rather than a numeric vector because there are non-numerical elements (such as a forward slash and tilde) in this variable.

## Explore your data graphically (Neonics)

9. Using `geom_freqpoly`, generate a plot of the number of studies conducted by publication year. 
```{r}
ggplot(Neonics) +
  geom_freqpoly(aes(x = Publication.Year))
```

10. Reproduce the same graph but now add a color aesthetic so that different Test.Location are displayed as different colors.
```{r}
ggplot(Neonics) +
  geom_freqpoly(aes(x = Publication.Year, color = Test.Location))
```

Interpret this graph. What are the most common test locations, and do they differ over time?

> Answer: The most common test locations alternate between the Lab and the Field natural, meaning they differ over time. 

11. Create a bar graph of Endpoint counts. What are the two most common end points, and how are they defined? Consult the ECOTOX_CodeAppendix for more information. 

```{r}
ggplot(Neonics, aes(x = Endpoint)) +
  geom_bar()
```

> Answer: The two most common end points are NOEL (No-observable-effect-level) and LOEL (Lowest-observable-effect-level); NOEL is defined as the highest dose (conentration) producing effects that were not significantly dfferent from responses of controls and LOEL is defined as the lowest dose (conentration) producing effects that were significantly different from responses of controls.

## Explore your data (Litter)

12. Determine the class of collectDate. Is it a date? If not, change to a date and confirm the new class of the variable. Using the `unique` function, determine which dates litter was sampled in August 2018. 
```{r}
class(Litter$collectDate)
Litter$collectDate <- as.Date(Litter$collectDate, format = "%Y-%m-%d") 
unique(Litter$collectDate)
```

13. Using the `unique` function, determine how many plots were sampled at Niwot Ridge. How is the information obtained from `unique` different from that obtained from `summary`?

```{r}
summary(Litter$plotID)
length(unique(Litter$plotID))
```

> Answer: 12 plots were sampled at Niwot Ridge. The information obtained from the 'unique' function is just the number of unique plot IDs, whereas the information obtained from the 'summary' is each plot ID and how many times they seperately were sampled. 

14. Create a bar graph of functionalGroup counts. This shows you what type of litter is collected at the Niwot Ridge sites. Notice that litter types are fairly equally distributed across the Niwot Ridge sites.
```{r}
ggplot(Litter, aes(x = functionalGroup)) +
  geom_bar()
```

15. Using `geom_boxplot` and `geom_violin`, create a boxplot and a violin plot of dryMass by functionalGroup. 
```{r}
ggplot(Litter) +
  geom_boxplot(aes(x = functionalGroup, y = dryMass))
ggplot(Litter) +
  geom_violin(aes(x = functionalGroup, y = dryMass))
```

Why is the boxplot a more effective visualization option than the violin plot in this case?

> Answer: The boxplot is a more effective visualization because it better represents the data distribution. The Violin plot can be misleading because certain types of litter have large dry mass outliers, although there is no way to tell that these data points are outliers in this type of visualization.

What type(s) of litter tend to have the highest biomass at these sites? 

> Answer: Needles tend to have the highest biomass at these sites. 